A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-04 inst/manuscript/manuscript.Rmd
abstract and introduction edits

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for design-based and model-based approaches to finite population spatial data. Then we use a simulation study and an analysis of real mercury concentration data to compare them numerically. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, model-based approaches tend to outperform design-based approaches, even for skewed data. This gap in performance is small when spatially balanced samples are used but large when non-spatially balanced samples are used.

@@

There are two general approaches for using data to make frequentist statistical inferences about a population: design-based and model-based. When data cannot be collected for all units in a population (population units), data are collected on a subset of the population units. This subset is called a sample. In the design-based approach, inferences about the underlying population are informed via a probabilistic process assigning some population units to the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process generating the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation].

@@

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
Though comparisons between design-based and model-based approaches to spatial data have been studied, no numerical comparison has been made between design-based approaches that incorporate spatial locations and model-based approaches. In this manuscript, we compare design-based approaches that incorporate spatial locations to model-based approaches for spatial data. We focus on finite populations, but these comparisons generalize to infinite populations as well. A finite population contains a finite number of population units; an example is lakes (treated as a whole with the lake centroid representing location) in the contiguous United States. An infinite population contains an infinite number of population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and compare several sampling and estimation procedures of the design-based and model-based approaches for finite population spatial data. In Section \ref{sec:numstudy}, we use a simulation approach to study the behavior and performance of both approaches. In Section \ref{application}, we use both approaches to analyze real data consisting of mercury concentration from lakes in the contiguous United States. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for the approaches and then use a simulation study and an analysis of real mercury concentration data to numerically compare them. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, the model-based approaches often outperform design-based approaches, even for skewed data. This gap in performance is small when spatially balanced samples are used but large when non-spatially balanced samples are used.

@@

There are two general approaches for using data to make frequentist statistical inferences about a population: design-based and model-based. When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units. This subset is called a sample. In the design-based approach, inferences about the underlying population are informed via a probabilistic process assigning some population units to be part of the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process generating the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation].

@@

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
Certainly comparisons between design-based and model-based approaches to spatial data have been studied. But no numerical comparison has been made between design-based approaches incorporating spatial information and design-based approaches. In this manuscript, we compare design-based approaches incorporating spatial information to model-based approaches for spatial data. We focus on finite populations, but these comparisons generalize to infinite populations as well. A finite population contains a finite number of population units; an example is lakes (treated as a whole with the lake centroid representing location) in the contiguous United States. An infinite population contains an infinite number of population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for the design-based and model-based approaches to finite population spatial data. In Section \ref{sec:numstudy}, we use a simulation study to compare the performance of the approaches in a variety of scenarios. In Section \ref{application}, we compare the performance of the approaches on real data that contains mercury concentration in lakes from the contiguous United States. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

### 2021-12-04 inst/manuscript/manuscript.tex
abstract and introduction edits

  \begin{abstract}
  The design-based and model-based approaches to frequentist statistical
  inference rest on fundamentally different foundations. In the
  design-based approach, inference depends on random sampling. In the
  model-based approach, inference depends on distributional assumptions.
  In this manuscript, we compare the approaches for finite population
  spatial data. We first provide relevant background for design-based and
  model-based approaches to finite population spatial data. Then we use a
  simulation study and an analysis of real mercury concentration data to
  compare them numerically. We find that sampling plans that incorporate
  spatial locations (spatially balanced samples) perform better than
  sampling plans ignoring spatial locations (non-spatially balanced
  samples), regardless of whether design-based or model-based approaches
  were used to analyze the data. We also find that within sampling plans,
  model-based approaches tend to outperform design-based approaches, even
  for skewed data. This gap in performance is small when spatially
  balanced samples are used but large when non-spatially balanced samples
  are used.
  \end{abstract}

@@

There are two general approaches for using data to make frequentist
statistical inferences about a population: design-based and model-based.
When data cannot be collected for all units in a population (population
units), data are collected on a subset of the population units. This
subset is called a sample. In the design-based approach, inferences
about the underlying population are informed via a probabilistic process
assigning some population units to the sample. Alternatively, in the
model-based approach, inferences are made from specific assumptions
about the underlying process generating the data. Each paradigm has a
deep historical context (Sterba, 2009) and its own set of benefits and
drawbacks (Hansen et al., 1983).

@@

Though comparisons between design-based and model-based approaches to
spatial data have been studied, no numerical comparison has been made
between design-based approaches that incorporate spatial locations and
model-based approaches. In this manuscript, we compare design-based
approaches that incorporate spatial locations to model-based approaches
for spatial data. We focus on finite populations, but these comparisons
generalize to infinite populations as well. A finite population contains
a finite number of population units; an example is lakes (treated as a
whole with the lake centroid representing location) in the contiguous
United States. An infinite population contains an infinite number of
population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and compare several sampling and
estimation procedures of the design-based and model-based approaches for
finite population spatial data. In Section \ref{sec:numstudy}, we use a
simulation approach to study the behavior and performance of both
approaches. In Section \ref{application}, we use both approaches to
analyze real data consisting of mercury concentration from lakes in the
contiguous United States. And in Section \ref{sec:discussion}, we end
with a discussion and provide directions for future research.

@@

The design-based approach can be used to select samples that are
``well-spread'' in space, or spatially balanced. Spatially balanced
samples are useful because parameter estimates from these samples tend
to vary less than parameter estimates from samples that are not
spatially balanced (Barabesi and Franceschi, 2011; Benedetti et al.,
2017; Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004). To
quantify the spatial balance of a sample, Stevens and Olsen (2004)
proposed loss metrics based on Voroni polygons. After the GRTS algorithm
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use the Generalized Random Tessellation Stratified
(GRTS) algorithm to select spatially balanced samples sampling because
the algorithm has several attractive properties. It accommodates finite
and infinite sampling frames. It accommodates equal, unequal, and
proportional (to size) inclusion probabilities. It accommodates legacy
(historical) sampling (Foster et al., 2017). It accommodates a minimum
distance between units in a sample. Lastly, it accommodates replacement
units in a sample, which are units that can be sampled in place of an
original unit that can no longer be sampled. The GRTS algorithm samples
from finite and infinite populations by utilizing a mapping between
two-dimensional and one-dimensional space. The units in the
two-dimensional sampling frame are divided into cells using a
hierarchical address. This hierarchical address is then used to map the
units from two-dimensional space to a one-dimensional line where each
unit's line length equals its inclusion probability. A systematic sample
is conducted on the line and linked back to a unit in two-dimensional
space, which results in the desired sample. Stevens and Olsen (2004)
provides further details.

>>>>

  \begin{abstract}
  The design-based and model-based approaches to frequentist statistical
  inference lie on fundamentally different foundations. In the
  design-based approach, inference depends on random sampling. In the
  model-based approach, inference depends on distributional assumptions.
  In this manuscript, we compare the approaches for finite population
  spatial data. We first provide relevant background for the approaches
  and then use a simulation study and an analysis of real mercury
  concentration data to numerically compare them. We find that sampling
  plans that incorporate spatial locations (spatially balanced samples)
  perform better than sampling plans ignoring spatial locations
  (non-spatially balanced samples), regardless of whether design-based or
  model-based approaches were used to analyze the data. We also find that
  within sampling plans, the model-based approaches often outperform
  design-based approaches, even for skewed data. This gap in performance
  is small when spatially balanced samples are used but large when
  non-spatially balanced samples are used.
  \end{abstract}

@@

There are two general approaches for using data to make frequentist
statistical inferences about a population: design-based and model-based.
When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units. This subset is called a sample. In the design-based approach,
inferences about the underlying population are informed via a
probabilistic process assigning some population units to be part of the
sample. Alternatively, in the model-based approach, inferences are made
from specific assumptions about the underlying process generating the
data. Each paradigm has a deep historical context (Sterba, 2009) and its
own set of benefits and drawbacks (Hansen et al., 1983).

@@

Certainly comparisons between design-based and model-based approaches to
spatial data have been studied. But no numerical comparison has been
made between design-based approaches incorporating spatial information
and design-based approaches. In this manuscript, we compare design-based
approaches incorporating spatial information to model-based approaches
for spatial data. We focus on finite populations, but these comparisons
generalize to infinite populations as well. A finite population contains
a finite number of population units; an example is lakes (treated as a
whole with the lake centroid representing location) in the contiguous
United States. An infinite population contains an infinite number of
population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
the design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:numstudy}, we use a simulation study to
compare the performance of the approaches in a variety of scenarios. In
Section \ref{application}, we compare the performance of the approaches
on real data that contains mercury concentration in lakes from the
contiguous United States. And in Section \ref{sec:discussion}, we end
with a discussion and provide directions for future research.

@@

The design-based approach can be used to select samples that are
``well-spread'' in space, or spatially balanced. Spatially balanced
samples are useful because parameter estimates from these samples tend
to vary less than parameter estimates from samples that are not
spatially balanced (Barabesi and Franceschi, 2011; Benedetti et al.,
2017; Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004). To
quantify the spatial balance of a sample, Stevens and Olsen (2004)
proposed loss metrics based on Voroni polygons. After the GRTS algorithm
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use the Generalized Random Tessellation Stratified
(GRTS) algorithm to select spatially balanced samples sampling because
the algorithm has several attractive properties. It accommodates finite
and infinite sampling frames, equal, unequal, and proportional (to size)
inclusion probabilities, legacy (historical) sampling (Foster et al.,
2017), a minimum distance between units in a sample, and replacement
units in a sample (which are units that can be sampled in place of an
original unit that can no longer be sampled). The GRTS algorithm samples
from finite and infinite populations by utilizing a mapping between
two-dimensional and one-dimensional space. The units in the
two-dimensional sampling frame are divided into cells using a
hierarchical address. This hierarchical address is then used to map the
units from two-dimensional space to a one-dimensional line where each
unit's line length equals its inclusion probability. A systematic sample
is conducted on the line and linked back to a unit in two-dimensional
space, which results in the desired sample. Stevens and Olsen (2004)
provides further details.
